\bf Visual diagnostics for constrained optimisation with application to guided tours

### 2020-11-02 paper/paper.tex
Update on Optmisation methods section

Visualisation is widely used in exploratory data analysis. Presenting information in a graphical format often allows people to uncover information they would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper is at the intersection of data visualization and optimization methods and aims to bring visualization tools into optimization problems to assist practitioners to better understand the performance of their optimization methods in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms \todo[inline]{(THERE ARE TWO THINGS HERE --> VISUAL METHODS + THE IMPLEMENTATION OF FINE TUNNING METHODS RIGHT ?}.
\todo[inline, color = green]{Add intro to fine tunning methods here too to set the scene}
The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained. The unconstrained problem can be formulated as a minimization (or maximization) problem such as
\(min_{x} f(x)\) where \(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with certain properties defined in \(f\in L^p\) . In this case typically solutions rely on gradient descent or ascent methods. While in the constrained problem additional restrictions are introduced via a set of functions that can be convex or non-convex \(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and hence the optimization problem in this case can be written as
\(min_{x} f(x)\) \emph{subject to} \(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex optimization methods includig linear and quadratic programming can be used.

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (REF) which is an exploratory data analysis tool that is defined to detect \emph{interesting structures} or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (REF). The target of the optimisation is thus to identify the most interesting low-dimensional views of the data given by a corresponding projection matrix. The most interesting structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space. \todo[inline]{CHALLENGES here that can also be set as general optimization challenges. However this depends on the sucess of index optimization, right?}
However, the optimization challenges encounter inthe projection pursuit guided tour problem are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are present in many areas of study and even though in some cases, analytical solutions can be found, the majority of optimization problems rely on numerical methods to find the optimal solution to a problem. In optimization, numerical methods follow interative approaches that aims at finding the optimum by progressively improving the current solution until a desirable accurary is achieved.

In this paper we focus on unconstrained optimization as defined in section XX and assume that it is not possible an analytical solution in closed form. That is, our problem consists of finding the mininum of function \(f \in L^p\) in the unconstrained \(\mathbb{R}\) space. In this scenario, gradient descent methods have taken the stage for many decades. In gradient descent methods the goal is to find the mininum, and the way to do so, is by taken steps proportional to the negative gradient of the function at the evaluation point \(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or the learning rate. The step size is the key in the sucess of the optimization problem here as inadequate step
sizes might lead to a failure in convergence. In order to solve this problem line minimization or line search methods are proposed (REF) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate \(x_{1}\) and then for each iteration a search direction is stablished \(S_k\) then the objective is to find
the value of \(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) (this step is called the line sear). Then move on to the next pont following \(x_{k+1} = x_k + \nu_kS_k\)
\todo[inline]{lconnection with derivative free methods}
\textbf{From Fletcher's book next paragraph --\textgreater{} clean up} Theoretical results have demonstrated the global convergence of the algorithm when the step-size \(\alpha_k\) is chosen by minimising \(f(x^{(j)} + \alpha* d^{(j)})\) \citep{curry1944method}. This is however rarely implemented in practice, either because of the high computational demand or because the solution does not exist \citep{fletcher2013practical}. A modified approach, in the case of maximisation, is to impose a mandatory increase in the objective function for each iteration: \(f^{(j+1)}> f^{(j)}\). Similarly, for minimisation we would require a decrease in function value at each step. This approach turns out to be efficient in practical problems, but global convergence cannot be guaranteed in this case.

@@

To define projection pursuit, we first need to establish the notation used. Let \(\mathbf{X}_{n \times p}\) be the data matrix, with \(n\) observations in \(p\) dimensions. A d-dimensional projection can be seen as a linear transformation \(T: \mathbb{R}^p \mapsto \mathbb{R}^d\), and defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where \(\mathbf{Y}_{n \times d}\) is the projected data and \(\mathbf{A}_{p\times d}\) is the projection matrix. Define \(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index function that maps the projected data \(\mathbf{Y}\) (corresponding to an associated projection matrix \(\mathbf{A}\)) onto an index value \(I\). This is commonly known as the projection pursuit index function, or just index function, and is used to measure the ``interestingness'' of a projection.

@@

As a general visualisation method, a tour produces animations of high dimensional data via rotations between low dimension planes. Different tour types choose these planes differently, for example, a grand tour\citep{cook2008grand} selects the planes randomly and a manual tour\citep{cook1997manual} gradually phases in and out one variable, to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses planes with the aid of projection pursuit. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs the geodesic interpolation between these planes to form a tour path.

@@

where \(I_d\) is the d-dimensional identity matrix and the constraint requires the projection bases \(\mathbf{A}\) to be an orthogonal matrices.

Several features of this optimisation are worth noticing. First of all, this is a constraint optimisation problem as the decision variables form the entries of a projection basis, which is required to be orthonormal. It is also likely that the objective function may not be differentiable for a constructed index function and in these cases, gradient-based methods may not work well. Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. Lastly, there is also one computational consideration: the optimisation procedure needs to be fast to compute since the tour animation is played in real-time.

@@

Simulated annealing (\texttt{search\_better\_random}) \citep[\citet{bertsimas1993simulated}]{kirkpatrick1983optimization} uses the same sampling process as \texttt{search\_better} but allow a probabilistic acceptance of a basis with lower index value based on the annealing \(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\) is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability \[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\] where \(I_{(\cdot)}\) denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maximum is presented. The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

\citet{cook1995grand} used a gradient ascent algorithm on the space of the projection bases. In gradient ascent, one first find the direction for improvement via computing the gradient information. In \texttt{search\_geodesic}, \(2n\) bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter \(\delta\). A geodesic is then constructed using the current basis and the one in \(2n\) bases with the highest index value. If the neighbourhood parameter \(\delta\) is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and works as the searching direction. The next step in gradient ascent is to conduct a line search to find the best improvement along the gradient direction and in \texttt{search\_geodesic}, this is replaced by optimising the index value along the geodesic direction over an 90 degree angle from \(-\pi/4\) to \(\pi/4\). The optima \(\mathbf{A}_{**}\) is returned for the current iteration if it meets the termination condition on percentage improvement. The procedure will also terminate if \(l_{\max}\) is reached. Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: \(\mathbf{A}\), 2) index values: \(I\), and 3) State: \(S\), which labels the observation with detailed stage in the optimisation. Possible values for \texttt{search\_better} and \texttt{search\_better\_random} includes \texttt{random\_search}, \texttt{new\_basis}, and \texttt{interpolation}. \texttt{search\_geodesic} has a wider variety that includes \texttt{new\_basis}, \texttt{direction\_search}, \texttt{best\_direction\_search}, \texttt{best\_line\_search}, and \texttt{interpolation}.

Multiple iterators are also needed to index the data collected at different levels. \(t\) is a unique identifier that prescribes the natural ordering of each observation; \(j\) is the counter for each search-and-interpolate iteration, which remains the same within one round and has an increment of one once a new round starts. \(l\) is the counter for each search/interpolation, which provides the information of how many basis the algorithm has searched before finding one to return. There are other parameters of interest, depends on the particular problem content and they are denoted as \emph{\(V_{p}\)}. Two most common examples include \(V_1 = \text{method}\), which tags the name of the algorithm used, and \(V_2 = \text{alpha}\), the neighbourhood parameter that controls the size in sampling candidate bases. A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is no output in iteration \(J + 1\) since the optimiser can't find a better basis and the algorithm terminates. In this notation, final basis found is \(A_T\) with the highest index value \(I_T\).

The data structure constructed above meets the tidy data principle \citep{wickham2014tidy} that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by the well-developed packages like dplyr\citep{dplyr} and ggplot2\citep{ggplot2}.

The construction of diagnostic plots uses the concept of grammar of graphic \citep{wickham2010layered} in ggplot2. In grammar of graphic, plots are not defined by its appearance (i.e.~boxplot, histogram, scatter plot etc) but by stacked layers. In the construction of diagnostic plots, there are multiple elements one may wants to emphasize and there's no single plot name that would meet the need, the stacked layers concept, on the other hand, allow information to be overlaid on the each other and one can build the plot from scratch as long as the variables have been stored in a dataset.

\hypertarget{check-how-hard-the-optimiser-is-working}{%
\subsection{Check how hard the optimiser is working}\label{check-how-hard-the-optimiser-is-working}}

A primary interest of diagnosing an optimiser is to study how it progressively finds its optimum. Directly plotting the index value across its natural order will cause the graph to be disproportional to the iteration since it usually takes longer for an optimiser to find a better basis towards the end. Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that can provide five points summary of each iteration. Other additional information not presented in the boxplot and then be added with new layers, for example text information on the number of points can be added at the bottom of each iteration and the position of basis returned by projection pursuit can be highlighted in point. Further, an option to switch between displaying points and boxplot geometry is helpful when the number of observation is small in one iteration and this is achieved via a \texttt{cutoff} parameter.

Figure \ref{fig:toy-search} shows a sample of the plot constructed using \texttt{search\_better} with different value on the parameter \texttt{max\_tries}. Comparing the returned index value of each iteration shows that the termination at \texttt{max\_tries\ =\ 25} is not sufficient for \texttt{search\_better} to explore the parameter space and a value of 500 is preferred over 25 in this context.

@@

\caption{A comparison of \texttt{search\_better} with different parameter value on \texttt{max\_tries}. A six variable dataset \texttt{boa6} is used with \texttt{holes} index on a 2D problem. A \texttt{max\_tries\ =\ 500} since it allows the optimiser to find better basis with higher index value at iteration six.}\label{fig:toy-search}
\end{figure}

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \ref{fig:toy-interp} presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decreases pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost cases is a deliberate construction of \texttt{search\_better\_random} where an inferior basis can be accepted in a probability manner so as to avoid trapping in the local maximum.

@@

\caption{1D projection on the 5 variable dataset \texttt{boa5} with two optimisers: \texttt{search\_better} and \texttt{search\_geodesic}. The yellow point corresponds to the theoretical best basis {[}0, 1, 0, 0, 0{]} with \texttt{V2} being the only non-normal variable in the dataset. The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \texttt{search\_geodesic} has been flipped positive to ensure that bases with same projected images but a sign difference are represented by the same point in the plot.}\label{fig:toy-pca}
\end{figure}

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases \(\mathbf{A}_{p \times d}\) lives on the surface of a \(p \times d\) dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space (d \textgreater{} 3), an easier alternative is to view the reduced space via some dimension reduction methods, i.e.~principal component analysis. To better perceive the the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package \texttt{geozoo} and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \ref{fig:toy-pca} plots the first two principal components of two search paths, one using \texttt{search\_better} and another uses \texttt{search\_geodesic}. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in \texttt{search\_better} is broader, controlled by \texttt{alpha} parameter, which is default to 0.5 while the directional search in \texttt{search\_geodesic}, controlled by \texttt{delta} with a default of 0.01, is so tiny that it can barely seen.

@@

\caption{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row view the space from a direction that is close to the one in PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.}\label{fig:toy-tour}
\end{figure}

Viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original \(p \times d\) dimension enables a more holistic stereoscopical view of the search. To view a high dimensional (\(d \ge 3\)) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compare to the PCA plot, the animated rotation (tour) display in Figure \ref{fig:toy-tour} gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and grand tour produces a sequence of angle that view the search from different direction. As an evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to one in Figure \ref{fig:toy-pca}.

@@

For a particular index function, the best algorithm to optimise relates to the character of the index and the data. If the index function is smooth and has single maximum, all of the three algorithms introduced above can find the maximum. When multiple optima are presented, \texttt{search\_better} may stuck at the local maximum and in the case where the index function is non-smooth, \texttt{search\_geodesic} may even fail to find the maximum. In this section, examples will be presented to outline how the diagnostic plots can be used to compare the performance of optimisers in different scenarios.

@@

Random variables with different structures has been simulated and the distribution of each is presented in Equation \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are normal distributed with zero mean and unit variance and \texttt{x2} to \texttt{x7} are mixtures of normal distributions with varied weights and locations. The mixture variables have been scaled to have an overall unit variance before running the projection pursuit.

@@

\hypertarget{monotonic}{%
\subsection{A problem of not monotonic}\label{monotonic}}

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have start the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds higher final index value with fewer steps as shown in the right panel of Figure \ref{fig:interruption}.

@@

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_better}. A comparison of the trace plot with and without the interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.}\label{fig:interruption}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the distance between the candidate basis and the current basis needs to be larger than 1e-3;
\item
  the percentage change of the index value need to be larger than 1e-5; and
\item
  the alpha parameter on itself need to be larger than 0.01
\end{enumerate}

>>>>

Visualisation is widely used in exploratory data analysis (REF). Presenting information in a graphical format often allows people to uncover information they would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper combines data visualization with optimization methods. It aims at bring visualization tools into optimization problems to assist users to better understand the performance of their optimization methods in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (REF). The unconstrained problem can be formulated as a minimization (or maximization) problem such as
\(min_{x} f(x)\) where \(f:\mathbb{R}^n \rightarrow \mathbb{R}\) is an objective function with certain properties defined in an \(L^p\) space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions are introduced via a set of functions that can be convex or non-convex: \(g_i:\mathbb{R}^n \rightarrow \mathbb{R}\) for \(i = 1, \ldots k\) and hence the problem can be written as
\(min_{x} f(x)\) \emph{subject to} \(g_i(x) \leq 0\). Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (REF) which is an exploratory data analysis tool that is defined to detect \emph{interesting structures} or features in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (REF). The target of the optimisation is to identify the most \emph{interesting} low-dimensional views of the data given by a corresponding projection matrix. The most \emph{interesting} structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (REF).

The optimization challenges encounter in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existance of multiple maxima (local and global), the tradeof between computational burden and proximity to the maxima, dealing with noisy objective functions, that might be non smooth and non diferentiable (REF). Those are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution to a problem. These numerical methods follow interative approaches that aim at finding the optimum by progressively improving the current solution until a desirable accurary is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existance of multiple maxima (local and global), the tradeoff between desirable accuracy and computational burden, as well as dealing with noisy objective functions. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

In this paper we focus on unconstrained optimization as defined in the introduction section and assume that it is not possible to find an analytical solution in closed form. That is, our problem consists of finding the mininum of a function \(f \in L^p\) in the unconstrained \(\mathbb{R}\) space. In this scenario, gradient descent methods have taken the stage for many decades now. In gradient descent methods the goal is to find the mininum, and the way to do so, is by taken steps proportional to the negative gradient of the function at the evaluation point \(x_{k+1} = x_k -\alpha_kS_k\) where \(\alpha_k\) is the step size or the learning rate (REF). The step size is the key in the sucess of the optimization problem as inadequate step
sizes might lead to a failure in convergence. In order to solve this problem line minimization or line search methods were proposed (REF) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate \(x_{1}\) and then for each iteration a search direction is stablished \(S_k\) then the objective is to find
the value of \(\alpha_k \in \mathbb{R}\) that minimizes \(f(x_k + \alpha_kS_k)\) with respect to \(\alpha_k\) (this step is called the line sear). Then move on to the next pont following \(x_{k+1} = x_k + \nu_kS_k\)
\todo[inline]{lconnection with derivative free methods}
\textbf{From Fletcher's book next paragraph --\textgreater{} clean up} Theoretical results have demonstrated the global convergence of the algorithm when the step-size \(\alpha_k\) is chosen by minimising \(f(x^{(j)} + \alpha* d^{(j)})\) \citep{curry1944method}. This is however rarely implemented in practice, either because of the high computational demand or because the solution does not exist \citep{fletcher2013practical}. A modified approach, in the case of maximisation, is to impose a mandatory increase in the objective function for each iteration: \(f^{(j+1)}> f^{(j)}\). Similarly, for minimisation we would require a decrease in function value at each step. This approach turns out to be efficient in practical problems, but global convergence cannot be guaranteed in this case.

@@

To define projection pursuit, we first need to establish the notation used. Let \(\mathbf{X}_{n \times p}\) be the data matrix, with \(n\) observations in \(p\) dimensions. A d-dimensional projection can be seen as a linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\), and defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where \(\mathbf{Y}_{n \times d}\) is the projected data and \(\mathbf{A}_{p\times d}\) is the projection matrix. Define \(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index function that maps the projected data \(\mathbf{Y}\) (corresponding to an associated projection matrix \(\mathbf{A}\)) onto an index value \(I\). This is commonly known as the projection pursuit index function, or just index function, and is used to measure the ``interestingness'' of a projection.

@@

As a general visualisation method, a tour produces animations of high dimensional data via rotations between low dimension planes. Different tour types choose these planes differently, for example, a grand tour \citep{cook2008grand} selects the planes randomly and a manual tour \citep{cook1997manual} gradually phases in and out one variable, to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses planes with the aid of projection pursuit. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs the geodesic interpolation between these planes to form a tour path.

@@

where \(I_d\) is the \(d\)-dimensional identity matrix and the constraint requires the projection bases \(\mathbf{A}\) to be orthogonal matrices.

Several features of this optimisation are worth noticing. First of all, this is a constrained optimisation problem as the decision variables form the entries of a projection basis, which is required to be orthonormal. It is also likely that the objective function may not be differentiable for a constructed index function and in these cases, gradient-based methods may not work well. Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. Lastly, there is also one computational consideration: the optimisation procedure needs to be fast to compute since the tour animation is played in real-time.

@@

Simulated annealing (\texttt{search\_better\_random}) \citep[\citet{bertsimas1993simulated}]{kirkpatrick1983optimization} uses the same sampling process as \texttt{search\_better} but allows a probabilistic acceptance of a basis with lower index value based on the annealing \(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\) is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability \[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\] where \(I_{(\cdot)}\) denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maxima are present. The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

\citet{cook1995grand} used a gradient ascent algorithm on the space of the projection bases. In gradient ascent, one first finds the direction for improvement via computing the gradient information. In \texttt{search\_geodesic}, \(2n\) bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter \(\delta\). A geodesic is then constructed using the current basis and the one in \(2n\) bases with the highest index value. If the neighbourhood parameter \(\delta\) is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and works as the searching direction. The next step in gradient ascent is to conduct a line search to find the best improvement along the gradient direction and in \texttt{search\_geodesic}, this is replaced by optimising the index value along the geodesic direction over an 90 degree angle from \(-\pi/4\) to \(\pi/4\). The optima \(\mathbf{A}_{**}\) is returned for the current iteration if it meets the termination condition on percentage improvement. The procedure will also terminate if \(l_{\max}\) is reached. Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: \(\mathbf{A}\), 2) index values: \(I\), and 3) State: \(S\), which labels the observation with detailed stage in the optimisation. Possible values for \texttt{search\_better} and \texttt{search\_better\_random} include \texttt{random\_search}, \texttt{new\_basis}, and \texttt{interpolation}. \texttt{search\_geodesic} has a wider variety that includes \texttt{new\_basis}, \texttt{direction\_search}, \texttt{best\_direction\_search}, \texttt{best\_line\_search}, and \texttt{interpolation}.

Multiple iterators are also needed to index the data collected at different levels. \(t\) is a unique identifier that prescribes the natural ordering of each observation; \(j\) is the counter for each search-and-interpolate iteration, which remains the same within one round and has an increment of one once a new round starts. \(l\) is the counter for each search/interpolation, which provides the information of how many bases the algorithm has searched before finding one to return. There are other parameters of interest, which depend on the particular problem content and they are denoted as \emph{\(V_{p}\)}. Two most common examples include \(V_1 = \text{method}\), which tags the name of the algorithm used, and \(V_2 = \text{alpha}\), the neighbourhood parameter that controls the size in sampling candidate bases. A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is no output in iteration \(J + 1\) since the fact that it is the last iteration means that the optimiser cannot find a better basis and the algorithm terminates. In this notation, the final basis found is \(A_T\) with the highest index value \(I_T\).

The data structure constructed above meets the tidy data principle \citep{wickham2014tidy} that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by well-developed packages such as dplyr\citep{dplyr} and ggplot2\citep{ggplot2}.

The construction of diagnostic plots uses the concept of grammar of graphics \citep{wickham2010layered} in ggplot2. In grammar of graphics, plots are not defined by their appearance (i.e.~boxplot, histogram, scatter plot etc) but by stacked layers. In the construction of diagnostic plots, there are multiple elements one may want to emphasize and there is no single plot name that would meet the need. The stacked layers concept, on the other hand, allows information to be overlaid on each other and one can build the plot from scratch as long as the variables have been stored in a dataset.

\hypertarget{checking-how-hard-the-optimiser-is-working}{%
\subsection{Checking how hard the optimiser is working}\label{checking-how-hard-the-optimiser-is-working}}

A primary interest of diagnosing an optimiser is to study how it progressively finds its optimum. Directly plotting the index value across its natural order will cause the graph to be disproportional to the iteration since it usually takes longer for an optimiser to find a better basis towards the end. Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that can provide five points summary of each iteration. Other additional information not presented in the boxplot can then be added with new layers, for example text information on the number of points can be added at the bottom of each iteration and the position of basis returned by projection pursuit can be highlighted in point. Further, an option to switch between displaying points and boxplot geometry is helpful when the number of observation is small in one iteration and this is achieved via a \texttt{cutoff} parameter.

Figure \ref{fig:toy-search} shows a sample of the plot constructed using \texttt{search\_better} with different values of the parameter \texttt{max\_tries}. Comparing the returned index value of each iteration shows that the termination at \texttt{max\_tries\ =\ 25} is not sufficient for \texttt{search\_better} to explore the parameter space and a value of 500 is preferred over 25 in this context.

@@

\caption{A comparison of \texttt{search\_better} with different values of the parameter \texttt{max\_tries}. A six-variable dataset \texttt{boa6} is used with \texttt{holes} index on a 2D problem. \texttt{max\_tries\ =\ 500} allows the optimiser to find better basis with higher index value at iteration six.}\label{fig:toy-search}
\end{figure}

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \ref{fig:toy-interp} presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decrease pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost case is a deliberate construction of \texttt{search\_better\_random} where an inferior basis can be accepted with some probability so as to avoid being trapped in a local maximum.

@@

\caption{1D projection on the 5-variable dataset \texttt{boa5} with two optimisers: \texttt{search\_better} and \texttt{search\_geodesic}. The yellow point corresponds to the theoretical best basis {[}0, 1, 0, 0, 0{]} with \texttt{V2} being the only non-normal variable in the dataset. The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \texttt{search\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.}\label{fig:toy-pca}
\end{figure}

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases \(\mathbf{A}_{p \times d}\) live on the surface of a \(p \times d\) dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space (\(d > 3\)), an easier alternative is to view the reduced space via some dimension reduction methods, i.e.~principal component analysis. To better perceive the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package \texttt{geozoo} and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \ref{fig:toy-pca} plots the first two principal components of two search paths, one using \texttt{search\_better} and the other using \texttt{search\_geodesic}. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in \texttt{search\_better} is broader, controlled by \texttt{alpha} parameter, which is default to 0.5 while the directional search in \texttt{search\_geodesic}, controlled by \texttt{delta} with a default of 0.01, is so tiny that it can barely be seen.

@@

\caption{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.}\label{fig:toy-tour}
\end{figure}

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original \(p \times d\) dimension enables a more holistic stereoscopic view of the search. To view a high dimensional (\(d \ge 3\)) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \ref{fig:toy-tour} gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \ref{fig:toy-pca}.

@@

For a particular index function, the best algorithm to optimise relates to the character of the index and the data. If the index function is smooth and has a single maximum, all of the three algorithms introduced above can find the maximum. When multiple optima are present, \texttt{search\_better} may get stuck at a local maximum and in the case where the index function is non-smooth, \texttt{search\_geodesic} may even fail to find the maximum. In this section, examples will be presented to outline how the diagnostic plots can be used to compare the performance of optimisers in different scenarios.

@@

Random variables with different structures have been simulated and the distribution of each is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are normal distributed with zero mean and unit variance and \texttt{x2} to \texttt{x7} are mixtures of normal distributions with varied weights and locations. The mixture variables have been scaled to have an overall unit variance before running the projection pursuit.

@@

\hypertarget{monotonic}{%
\subsection{A problem of non-monotonicity}\label{monotonic}}

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have started the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds a higher final index value with fewer steps as shown in the right panel of Figure \ref{fig:interruption}.

@@

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_better}. A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.}\label{fig:interruption}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  the distance between the candidate basis and the current basis needs to be larger than 1e-3;
\item
  the percentage change of the index value need to be larger than 1e-5; and
\item
  the alpha parameter on itself needs to be larger than 0.01
\end{enumerate}
